{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2024-03-13 paper/main.tex
first revision of the manuscript following the reviewers' comments

@@

Our main contribution is an application of these ideas to direct reciprocity, which is one of the key mechanisms to explain why unrelated individuals might cooperate~\citep{nowak:Science:2006}. 
According to this mechanism, cooperation pays if it makes the interaction partner more cooperative in future. 
To describe which strategies are most effective, the previous theoretical literature assumes that the evolutionary dynamics are driven by the players' expected payoffs~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,van-segbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022}.
To the extent that strategies are learned (not inherited), this assumption seems to impose rather stringent requirements on the individuals' cognitive abilities. 
In the most extreme case, this would require individuals to remember (or compute) their payoffs against all population members, for all possible ways in which their repeated games may unfold. 
This assumption introduces a curious inconsistency in how these models represent an individual's cognitive abilities. 
For playing their games, individuals are often assumed to only recall the outcome of the very last round. 
Yet to update their strategies, individuals are implicitly assumed to have a record of the outcome of all rounds, across all interaction partners.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{static/donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    The leftmost panels give a schematic overview of the two main scenarios we compare. 
    The two scenarios differ in how many past interactions individuals take into account when updating their strategy. 
    In the scenario with perfect payoff memory, individuals consider all their past interactions (against all population members, and taking every turn of each repeated game into account). 
    In the scenario with limited payoff memory, individuals only consider their very last interaction (against one specific population member, taking into account only one round of the repeated game). 
    The four panels  on the right side depict the outcome of evolutionary simulations for repeated games with either a low or a high benefit of cooperation. 
    Colors represent how often the respective region of the strategy space is visited over time. 
    In all four panels, two regions are visited particularly often. 
    One region corresponds to a neighborhood of \alld{} with $p\approx q\!\approx\!0$ (lower left corner). 
    The other region corresponds to a strip of conditionally cooperative strategies with $p\!\approx\! 1$ and $q$ satisfying the constraints \eqref{Eq:PerfectMemory} and \eqref{Eq:LimitedMemory}, respectively (lower right corner). 
    The resulting average cooperation rate depends on which of these two neighborhoods is visited more often. 
    Simulations are run for $T\!=\!10^7$ time steps, using a cost $c\!=\!1$, a continuation probability of $\delta\!=\!0.999$ and a selection strength of $\beta\!=\!1$, in a population of size $N\!=\!100$.}
\label{fig:expected_and_stochastic_for_donation}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=.75\textwidth]{static/cooperation_rate_over_b_and_beta.pdf}
    \caption{{\bf Evolution of direct reciprocity for different parameter values.} 
    To explore the robustness of our results, we have run simulations for different benefit values (left panels, {\it a}), and for different selection strengths (right panels, {\it b}). 
    In each case, we record the resulting average cooperation rate over the entire simulation (upper panels). 
    In addition we record the individuals' average generosity. 
    Here, we only take into account those residents with $p\! \approx\! 1$ and we compute
    the average of their cooperation probability~$q$. 
    These simulations suggest that perfect payoff memory consistently leads to more cooperation and more generosity. 
    Unless explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. 
    Simulations are run for $T\!=\!5\times 10^7$ time steps for each parameter combination.}
    \label{fig:cooperation_rate_over_benefit_and_beta}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{static/more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different payoff memories.}
  We vary how much information individuals take into account when updating their strategies. 
  From left to right, we consider the following cases.
  ({\it i}) Updating occurs based on expected payoffs (perfect memory), 
  ({\it ii}) it occurs based on the last round of one interaction (limited memory), 
  ({\it iii}) based on the last round of two interactions,
  ({\it iv}) based on the last two rounds of one interaction,
  ({\it v}) based on the last two rounds of two interactions, and
  ({\it vi}) based on the average payoff of one interaction.
  Again, simulations are run either for a comparably low benefit of cooperation~($b/c\!=\!3$), or for a high benefit~($b/c\!=\!10$). 
  We observe that perfect memory always yields the highest cooperation rate. 
  However, when individuals take into account at least two past interactions -- cases ({\it iii}) to ({\it vi}) -- evolving cooperation rates are close to this optimum. 
  Baseline parameters are the same as in \FigDependenceParameters.}
\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}

>>>>

%% RESULTS: Snowdrift Game
{\bf Beyond the donation game.}
So far, we have discussed the donation game, which is the most common
example of social dilemmas. There are several other examples, which include the
snowdrift game~\cite{doebeli:Ecology:2005}, the stag hunt
game~\cite{pacheco:ProcB:2009}, and the public goods
game~\cite{pinheiro:PLoSCompBio:2014}. However, the pattern we observe is more
general and still holds for other social dilemmas. To demonstrate this, we
consider the snowdrift game. In the snowdrift game, individuals can again
cooperate or defect. However, now if both cooperate, they have the benefit \(b\)
of getting home while sharing the labor \(c\). Thus for mutual cooperation, they
receive \(b - \frac{c}{2}\). However, if one defects, they get \(b\) while
the cooperator gets \(b - c\). By contrast to the donation game, now
the best action depends on the co-player: to defect if the other cooperates, but
to cooperate if the other defects.

For most scenarios, we can again derive an analytical condition for when
cooperation is stochastically stable. Here we focus only on the case of perfect
and limited memory. The details can be found in the electronic supplementary
material. In the perfect memory case, we find that cooperation is stochastically
stable when \(q < 1 - \frac{c}{2 \delta b}\), and interestingly for the limited
memory case \(q < 1 - \frac{1}{2 \delta}\), which is the same as in
Eq.~\eqref{Eq:LimitedMemory}. This is because in this analysis, the ordering of
the payoffs does not matter too much; what matters most is the following.

Apart from the analytical analysis, we evaluate our results with simulations,
and we do this for all the updating payoff cases considered in this manuscript.
What we find is that the qualitative results still hold for the snowdrift game.
Also, in that case, we observe that perfect memory leads to systematically
higher cooperation rates (Supplementary Figures~8, 9). However, now less memory
is needed to approximate the classical limit.

@@

Our main contribution in this work is the application of these ideas to direct reciprocity, which is one of the key mechanisms to explain why unrelated individuals might cooperate~\citep{nowak:Science:2006}. We discuss the main contributions of our work in the next section.

%%%%%%%%%%%%
%%  Conclusion %%
%%%%%%%%%%%%

\section{Conclusion}\label{section:Conclusion}

According to the mechanism of direct reciprocity, cooperation pays if it makes the interaction partner more cooperative in future. 
To describe which strategies are most effective, the previous theoretical literature assumes that the evolutionary dynamics are driven by the players' expected payoffs~\citep{brauchli:JTB:1999,brandt:JTB:2006,ohtsuki:JTB:2007b,szolnoki:pre:2009b,imhof2010stochastic,van-segbroeck:prl:2012,grujic:jtb:2012,Martinez2012,stewart:pnas:2013,pinheiro:PLoSCB:2014,stewart:games:2015,Baek2016,McAvoy:ProcA:2019,glynatsi:SCR:2020,Schmid:PlosCB:2022,Murase:SciRep:2022}.
To the extent that strategies are learned (not inherited), this assumption seems to impose rather stringent requirements on the individuals' cognitive abilities. 
In the most extreme case, this would require individuals to remember (or compute) their payoffs against all population members, for all possible ways in which their repeated games may unfold. 
This assumption introduces a curious inconsistency in how these models represent an individual's cognitive abilities. 
For playing their games, individuals are often assumed to only recall the outcome of the very last round. 
Yet to update their strategies, individuals are implicitly assumed to have a record of the outcome of all rounds, across all interaction partners.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{static/donation_expected_last_round_summary_results.pdf}
    \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
    The leftmost panels give a schematic overview of the two main scenarios we compare. 
    The two scenarios differ in how many past interactions individuals take into account when updating their strategy. 
    In the scenario with perfect payoff memory, individuals consider all their past interactions (against all population members, and taking every turn of each repeated game into account). 
    In the scenario with limited payoff memory, individuals only consider their very last interaction (against one specific population member, taking into account only one round of the repeated game). 
    The four panels  on the right side depict the outcome of evolutionary simulations for repeated games with either a low or a high benefit of cooperation. 
    Colors represent how often the respective region of the strategy space is visited over time. 
    In all four panels, two regions are visited particularly often. 
    One region corresponds to a neighborhood of \alld{} with $p\approx q\!\approx\!0$ (lower left corner). 
    The other region corresponds to a strip of conditionally cooperative strategies with $p\!\approx\! 1$ and $q$ satisfying the constraints \eqref{Eq:PerfectMemory} and \eqref{Eq:LimitedMemory}, respectively (lower right corner). 
    The resulting average cooperation rate depends on which of these two neighborhoods is visited more often.
    Given the stochastic nature of the evolutionary process, we run long-time simulations to thoroughly explore its behavior and capture significant trends.
    Simulations are run for $T\!=\!10^7$ time steps, using a cost $c\!=\!1$, a continuation probability of $\delta\!=\!0.999$ and a selection strength of $\beta\!=\!1$, in a population of size $N\!=\!100$.}
\label{fig:expected_and_stochastic_for_donation}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=.75\textwidth]{static/cooperation_rate_over_b_and_beta.pdf}
    \caption{{\bf Evolution of direct reciprocity for different parameter values.} 
    To explore the robustness of our results, we have run simulations for different benefit values (left panels, {\it a}), and for different selection strengths (right panels, {\it b}). 
    In each case, we record the resulting average cooperation rate over the entire simulation (upper panels). 
    In addition we record the individuals' average generosity. 
    Here, we only take into account those residents with $p\! \approx\! 1$ and we compute
    the average of their cooperation probability~$q$. 
    These simulations suggest that perfect payoff memory consistently leads to more cooperation and more generosity. 
    Unless explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$.
    Each point represents a single simulation run. Simulations are run for $T\!=\!5\times 10^7$ time steps for each parameter combination. Each point is a sinlge simualtion run.}
    \label{fig:cooperation_rate_over_benefit_and_beta}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{static/more_memory_summary_results.pdf}
  \caption{{\bf Average cooperation rates for different payoff memories.}
  We vary how much information individuals take into account when updating their strategies. 
  From left to right, we consider the following cases.
  ({\it i}) Updating occurs based on expected payoffs (perfect memory), 
  ({\it ii}) it occurs based on the last round of one interaction (limited memory), 
  ({\it iii}) based on the last round of two interactions,
  ({\it iv}) based on the last two rounds of one interaction,
  ({\it v}) based on the last two rounds of two interactions, and
  ({\it vi}) based on the average payoff of one interaction.
  Again, simulations are run either for a comparably low benefit of cooperation~($b/c\!=\!3$), or for a high benefit~($b/c\!=\!10$). 
  We observe that perfect memory always yields the highest cooperation rate. 
  However, when individuals take into account at least two past interactions -- cases ({\it iii}) to ({\it vi}) -- evolving cooperation rates are close to this optimum. 
  Simulations are run for $T\!=\!10^7$ time steps, using a cost $c\!=\!1$, a continuation probability of $\delta\!=\!0.999$ and a selection strength of $\beta\!=\!1$, in a population of size $N\!=\!100$.}
\label{fig:cooperation_rate_all_updating_payoffs}
\end{figure}
